---
title: "Getting Started with macrox"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with macrox}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

macrox records every step of a PDF table extraction workflow as a YAML macro and replays it automatically on new files.  Think of it as Power Query for PDFs — interactive exploration that compiles into a reproducible pipeline.

---

## Installation

```{r install}
remotes::install_github("cathalbyrnegit/macrox")

# Optional but recommended
install.packages(c(
  "magick",      # page annotation and image cropping
  "ellmer",      # LLM extraction
  "validate",    # data quality rules
  "shinyAce",    # YAML editor in app
  "shinyFiles",  # filesystem browser in app
  "shinyjs",     # JS helpers in app
  "writexl",     # Excel export
  "jsonlite"     # JSON export
))
```

---

## Quick start

```{r quickstart}
library(macrox)

# Open a session
sess <- mx_session("report_2024.pdf")

# Scan for tables
sess |> detect_tables(pages = 1:20)

# Draw an area, extract, clean
area <- sess |> locate_area(page = 8)
sess |> select_table("monthly", page = 8, area = area, method = "bbox")
sess |> rename_columns("monthly", c(Month = "month", Total = "total"))
sess |> cast_types("monthly", c(total = "integer"))
sess |> filter_rows("monthly", exclude_where = "month == 'Total'")

# Save the macro
sess |> save_macro("my_report")

# Replay next year
mx_replay("report_2025.pdf", macro = "my_report")
```

---

## Standalone app

```{r app}
mx_app()           # system browser
mx_app("dialog")   # RStudio dialog
mx_app("pane")     # Viewer pane
```

The app provides a full extraction workflow without writing any R code.  Select a PDF with the filesystem browser, draw a bounding box on the page image, configure the extraction method, and hit **Extract**.  All steps are recorded and can be exported as a YAML macro or downloaded as CSV / Excel.

**Tabs:**

| Tab | Purpose |
|---|---|
| **Extract** | Full PDF viewer, draw-to-select area, method picker (bbox / lattice / stream / docling / llm), live preview. `←` / `→` navigate pages. |
| **Tables** | One tab per extracted table. Rename, cast types, filter rows (GUI), validate rules, column stats. |
| **Items** | Extract single metadata fields (invoice numbers, dates, totals) via LLM or GLiNER2. |
| **Struct** | Extract structured records (multiple fields, multiple rows) via GLiNER2. |
| **Steps** | Recorded step badges. Remove steps, edit the YAML macro directly with syntax highlighting. |
| **Replay** | Run a saved macro against a new PDF. Results load into Tables. |
| **Batch** | Run one macro across multiple PDFs. Download as Excel or zip of CSVs. |
| **Export** | Download CSV, Excel, JSON, or the YAML macro. |

---

## Extraction methods

macrox ships five extraction engines.  They can be mixed freely within the same session and macro.

### bbox — word-position engine (recommended default)

Uses `pdftools::pdf_data()` word-level bounding boxes.  No Java required.  Best for PDFs with charts on the same page, missing grid lines, or numbers running together.

```{r bbox}
area <- sess |> locate_area(page = 16)

sess |> select_table("breed_sire", page = 16, area = area, method = "bbox")

# Tune when columns merge or split
sess |> select_table("monthly", page = 27, method = "bbox",
  area    = c(top = 80, left = 30, bottom = 500, right = 750),
  col_gap = 20,    # minimum horizontal gap (pts) between columns
  row_tol = 3)     # vertical tolerance for grouping words into rows
```

### lattice / stream — tabulapdf engines

For clean PDFs with visible grid lines (`lattice`) or whitespace-aligned columns (`stream`).  Requires Java and the `tabulapdf` package (listed in Suggests — not required for other methods).

```{r tabulapdf}
sess |> select_table("calf_monthly", page = 8, method = "lattice")
sess |> select_table("calf_stream",  page = 9, method = "stream")
```

### docling — ML layout engine (IBM Research)

Uses Docling's deep learning layout models to detect and extract tables.  Converts scanned PDFs and handles complex multi-column layouts where positional methods fail.  Runs fully offline after a one-time model download (~2-3 GB).  No API key required.

```{r docling-setup}
# One-time setup — installs docling into a Python virtualenv
setup_docling()
# Restart R before first use
```

```{r docling}
# Scan for tables using Docling's ML detection (single-pass, cached)
sess |> detect_tables(pages = 1:20, method = "docling")

# Extract a specific table
sess |> select_table_docling("financials", page = 10, table_index = 1)

# Release the document cache when done
close_docling()
```

Docling caches converted pages in memory, so repeated calls for different tables on the same page are fast.  Use `close_docling()` to release the cache.

### llm — LLM engine via ellmer

Sends a rendered page image to a large language model.  Best for multi-level spanning headers, scanned PDFs, and any layout where positional extraction fails.  Requires the `ellmer` package and a provider API key in `.Renviron`.

```{r llm}
# Pass any ellmer chat object directly
chat <- ellmer::chat_anthropic()
sess |> select_table_llm("stillborn", page = 40, area = area, chat = chat)

# Or specify provider/model inline (chat object created automatically)
sess |> select_table_llm("stillborn", page = 40, area = area,
                          provider = "anthropic")

# Predefined schema — more accurate and consistent across replays
sess |> select_table_llm("breed_sire", page = 16, area = area,
  chat     = chat,
  schema   = c(Breed = "character", Male = "integer",
               Female = "integer", Total = "integer"),
  prompt   = "Flatten the two-row header with _ separators.")

# Refine schema without re-drawing the area
sess |> update_llm_schema("breed_sire",
  schema = c(Breed = "character", Male = "integer", Total = "integer"))
```

**Using `chat =`** — pass any [ellmer](https://ellmer.tidyverse.org/) chat
object for full control over model parameters, system prompts, and provider
configuration:

```{r llm-chat}
chat <- ellmer::chat_openai(model = "gpt-4o")
chat <- ellmer::chat_google_gemini(model = "gemini-2.0-flash")
chat <- ellmer::chat_ollama(model = "llava")

# Same chat object works for tables and items
sess |> select_table_llm("tbl", page = 1, chat = chat)
sess |> select_item("ref", prompt = "Reference number", chat = chat)
```

**Using `provider =`** — shorthand when defaults are fine.  `provider` can be
the name of any ellmer chat constructor without the `chat_` prefix — e.g.
`"anthropic"`, `"openai"`, `"google_gemini"`, `"openrouter"`, `"groq"`,
`"mistral"`, `"deepseek"`, `"ollama"`, or `"openai_compatible"`.

```{r llm-compatible}
sess |> select_table_llm("tbl", page = 5,
  provider = "openai_compatible",
  model    = "llama3.2-vision",
  base_url = "http://localhost:1234/v1")

sess |> select_table_llm("tbl", page = 5,
  provider = "openrouter",
  model    = "anthropic/claude-opus-4.5")
```

**API keys** — set in `.Renviron`, using whatever environment variable the
chosen `ellmer::chat_<provider>()` expects:
```
ANTHROPIC_API_KEY=sk-ant-...
OPENAI_API_KEY=sk-...
GEMINI_API_KEY=AI...
OPENROUTER_API_KEY=sk-or-...
```

---

## Multi-page tables

When the same table spans several consecutive PDF pages, `stack_pages()` extracts from each page and row-binds the results.  Repeated headers are consumed automatically.

```{r stack-pages}
sess |> stack_pages("breed_all", pages = 10:14,
  area   = c(top = 80, left = 30, bottom = 700, right = 550),
  method = "bbox")

# When pages 2+ don't repeat the header row:
sess |> stack_pages("breed_all", pages = 10:14,
  method       = "bbox",
  header_match = FALSE)
```

---

## Extracting metadata items

`select_item()` extracts a single field from the document — invoice numbers, dates, totals, report titles — anything that isn't a table row.  Results sit alongside tables in `sess$items` and are included in `export_json()`.

```{r items}
# Text-based (default): sends all page text to the LLM — fast and cheap
sess |> select_item("invoice_number",
  prompt = "Extract the invoice ID or reference number.")

sess |> select_item("invoice_date",
  prompt = "The billing issue date.",
  cast   = "date:%d/%m/%Y")

sess |> select_item("total_amount",
  prompt = "The grand total payable.",
  cast   = "numeric")

# Image-based: render a specific page (better for scanned or visual PDFs)
sess |> select_item("stamp_ref", page = 1L,
  prompt = "The approval stamp reference in the top-right corner.",
  area   = c(top = 20, left = 500, bottom = 100, right = 780))

# Review all items
sess |> show_items()
```

### GLiNER2 backend — local, no API key

The GLiNER2 backend extracts fields using a local NLP model.  Runs on CPU, no API key needed.

```{r items-gliner}
# One-time setup
setup_gliner()

# Single field
sess |> select_item("invoice_number",
  prompt  = "Invoice ID or reference number",
  backend = "gliner")

# Batch extraction — multiple fields in one model pass (much faster)
sess |> select_items_batch(c(
  invoice_no = "Invoice ID or reference number",
  total      = "Grand total amount payable",
  due_date   = "Payment due date"
), cast = c(total = "numeric", due_date = "date:%d/%m/%Y"))
```

### Structured record extraction

`select_struct()` extracts multiple structured records — like a table, but from unstructured text using GLiNER2.

```{r struct}
sess |> select_struct("line_items",
  entity = "LineItem",
  fields = c(
    description = "Product or service description",
    quantity    = "Number of units",
    unit_price  = "Price per unit"
  ),
  list_fields = character(0),
  page = 1L)

# Convert to a data frame (with optional confidence filtering)
sess |> struct_to_df("line_items", min_confidence = 0.5)

# The result is now in sess$tables$line_items — clean, rename, export as usual
sess |> preview("line_items")
```

---

## Cleaning tables

```{r cleaning}
sess |> rename_columns("monthly", c(
  Month = "month", Male = "male_count", Female = "female_count", Total = "total"
))

sess |> cast_types("monthly", c(
  male_count   = "integer",
  female_count = "integer",
  total        = "integer"
  # also: "numeric", "character", "date:%d/%m/%Y", "date:%Y-%m-%d"
))

sess |> filter_rows("monthly", exclude_where = "month == 'Total'")
```

### Forward-fill and number cleaning

Government PDFs often have merged header cells that leave blanks in rows below.  `fill_down()` propagates the last non-blank value downward.

```{r fill-clean}
# Forward-fill blank cells in specific columns
sess |> fill_down("monthly", cols = "region")

# Forward-fill all columns
sess |> fill_down("monthly")

# Strip currency symbols, thousands separators, parenthetical negatives
sess |> clean_numbers("monthly", cols = c("male_count", "female_count"))

# clean_numbers() with cols = NULL targets all non-numeric columns
sess |> clean_numbers("monthly")
```

### Suggesting column types

`suggest_schema()` inspects values and proposes a `cast_types()` call:

```{r suggest}
sess |> suggest_schema("monthly")
#> Suggested schema for 'monthly'
#>   month         ->  character
#>   male_count    ->  integer
#>   female_count  ->  integer
#>   total         ->  integer
```

---

## Deriving and combining tables

These steps are all recorded in the macro and replay automatically.

```{r derive}
# Add a derived column (expression evaluated against the data frame)
sess |> add_column("monthly", "year",  expr = "2024L")
sess |> add_column("monthly", "ratio", expr = "male_count / total")

# Row-bind tables with the same schema
sess |> stack_tables("all_years",
  tables = c("calves_2022", "calves_2023", "calves_2024"))

# Optional: pad missing columns with NA when schemas differ slightly
sess |> stack_tables("all_years",
  tables = c("calves_2022", "calves_2023"),
  .fill  = TRUE)

# Join two tables
sess |> merge_tables("combined",
  left  = "monthly",
  right = "county_totals",
  by    = "month")
```

---

## Data validation

`validate_table()` uses the `validate` package DSL.  Rules are stored in the macro and re-run automatically on every replay — catching data drift silently.

```{r validate}
sess |> validate_table("monthly", rules = c(
  twelve_rows    = "nrow(.) == 12",
  no_na_month    = "!any(is.na(month))",
  positive_males = "all(male_count > 0, na.rm = TRUE)",
  totals_sum     = "all(abs(male_count + female_count - total) < 2, na.rm = TRUE)"
))
# ✔ All 4 validation rules passed for 'monthly'

# Review all validation results
sess |> show_validations()

# Strict mode — abort replay on failure (good for production pipelines)
sess |> validate_table("monthly", rules = c(...), strict = TRUE)
```

---

## Reviewing and editing steps

```{r steps}
sess |> show_steps()
# * [1]  select_table   → monthly       | page 8 [bbox]
# * [2]  rename_columns → monthly       | 4 renames
# * [3]  cast_types     → monthly       | 4 casts
# * [4]  filter_rows    → monthly       | exclude: month == 'Total'
# * [5]  add_column     → monthly       | year = 2024L
# * [6]  validate_table → monthly       | 4 rules
# * [7]  select_item    → invoice_date  | cast: date

sess |> remove_step(3)           # remove by index; clears table if extraction step
sess |> preview_all()            # inspect all tables before saving
```

---

## Saving and replaying

```{r save-replay}
# Save macro to disk
sess |> save_macro("my_report", path = "inst/macros/")

# Replay against a new file — returns named list of data frames
tables <- mx_replay("report_2025.pdf", macro = "my_report")

# Batch replay across multiple files
files   <- list.files("reports/", pattern = "\\.pdf$", full.names = TRUE)
results <- mx_replay_batch(files, macro = "my_report")
```

---

## Comparing runs

```{r diff}
diff <- diff_replay(
  "report_2024.pdf",
  "report_2025.pdf",
  macro = "my_report"
)
print(diff)
# macrox diff
#   Reference: report_2024.pdf
#   New:       report_2025.pdf
#
# ✔ monthly        [unchanged]
# ~ breed_by_sire  [changed] (rows: 17→18, +cols: HE, 2 cell changes)
# ✔ calf_monthly   [unchanged]
```

---

## Exporting data

```{r export}
# CSV files — one per table
sess |> export_csv(dir = "output/")

# Single Excel workbook — one sheet per table
sess |> export_excel("output/report_tables.xlsx")

# JSON — items (metadata) + tables (row arrays) in one payload
json <- sess |> export_json()                       # returns string
sess |> export_json(path = "output/report.json")    # writes file
sess |> export_json(include_steps = TRUE)           # include macro steps
```

The JSON output structure:

```json
{
  "source": "report_2024.pdf",
  "extracted": "2026-05-01T14:32:00",
  "items": {
    "invoice_number": "INV-9821",
    "total_amount": 1234.56
  },
  "tables": {
    "monthly": [
      { "month": "January", "total": 4700 },
      { "month": "February", "total": 11270 }
    ]
  }
}
```

---

## Agent integration

`test_extraction()` is the **R-Core Judge** for agentic workflows.  It runs one extraction pass and returns a structured evaluation report — enabling an MCP agent to self-correct its parameters in a feedback loop.

```{r agent}
# Single test pass — no session needed
report <- test_extraction(
  file        = "invoice.pdf",
  page        = 1L,
  area        = c(top = 200, left = 30, bottom = 600, right = 780),
  method      = "bbox",
  header_rows = 1L
)

report$status
#> [1] "needs_alignment"

report$metrics
#> $rows_extracted      [1] 12
#> $unnamed_headers     [1] "V1" "V2"
#> $header_spill        [1] TRUE
#> $empty_columns       list()

report$guidance
#> [1] "Header text detected in data rows — increase header_rows by 1. |
#>       Unnamed columns (V1, V2) — adjust area coords or increase header_rows."
```

The return value serialises directly to JSON via `jsonlite::toJSON()`, making it straightforward to wire into an MCP `test_parameters` tool.  The agent loop runs `test_extraction()` repeatedly — adjusting `area`, `header_rows`, and `method` based on `$guidance` — until `status == "success"`, then calls `save_macro()` to lock the verified parameters.

Pre-flight validation before replay:

```{r agent-validate}
# Validate a macro file without running it
result <- validate_macro("report_2025.pdf", macro = "my_report")
print(result)
# ✔ Macro valid
#   ✔ [1] select_table / monthly
#   ✔ [2] rename_columns / monthly
```

---

## The YAML macro

All step types serialise cleanly.  The full extraction parameters are stored so replay is exact.

```yaml
macro:
  name: my_report
  created: '2026-05-01 14:32'
  source: report_2024.pdf
  n_steps: 10

steps:
  - step: select_table
    label: monthly
    page: 8
    method: bbox
    area: [80, 30, 500, 750]
    header_rows: 1

  - step: select_table_docling
    label: financials
    page: 10
    table_index: 1

  - step: select_table_llm
    label: breed_sire
    page: 16
    area: [60, 25, 620, 790]
    provider: anthropic
    model: claude-opus-4-8
    schema:
      Breed: character
      Male: integer
      Total: integer
    prompt: "Flatten the two-row header."

  - step: select_item
    label: report_date
    prompt: "The publication date of this report."
    cast: "date:%B %Y"
    backend: llm
    provider: anthropic

  - step: select_struct
    label: line_items
    entity: LineItem
    fields:
      description: "Product or service description"
      quantity: "Number of units"
    page: 1

  - step: rename_columns
    table: monthly
    mapping:
      Month: month
      Total: total

  - step: cast_types
    table: monthly
    types:
      total: integer

  - step: fill_down
    table: monthly
    cols: [region]

  - step: clean_numbers
    table: monthly
    cols: [total]

  - step: validate_table
    table: monthly
    strict: false
    rules:
      twelve_rows: "nrow(.) == 12"
      no_na:       "!any(is.na(month))"
```

---

## Shiny module

Embed macrox in an existing Shiny app:

```{r module}
# UI
bslib::accordion_panel(
  "PDF Import",
  icon = shiny::icon("file-pdf"),
  macrox::macrox_ui(ns("pdf_import"))
)

# Server
pdf_result <- macrox::macrox_server("pdf_import")

observe({
  tables <- pdf_result$tables()
  req(length(tables) > 0)
  for (nm in names(tables)) your_upload_fn(tables[[nm]], target = nm)
})
```

---

## Choosing an extraction method

| Situation | Recommended |
|---|---|
| Clean PDF with visible grid lines | `lattice` |
| Whitespace-aligned columns, no grid | `stream` |
| Charts and tables on the same page | `bbox` + `area =` |
| Numbers running together | `bbox` + `col_gap =` |
| Scanned PDF, no selectable text | `docling` or `llm` |
| Complex layout, offline | `docling` |
| Multi-level / spanning headers | `llm` with `schema =` |
| Complex irregular layout | `llm` |
| No API key / offline | `bbox`, tabulapdf, or `docling` |
| Single metadata fields | `select_item()` with `llm` or `gliner` |
| Structured records from text | `select_struct()` with GLiNER2 |

---

## Dependencies

| Package | Role | Required |
|---|---|---|
| `pdftools` | Word positions (`bbox`), page text, rendering | Yes |
| `stringdist` | Fuzzy caption matching | Yes |
| `yaml` | Macro read/write | Yes |
| `cli` | Console output | Yes |
| `rlang` | `%\|\|%` operator | Yes |
| `tabulapdf` | `lattice` / `stream` (needs Java) | Suggests |
| `reticulate` | Python bridge for Docling and GLiNER2 | Suggests |
| `magick` | Page annotation; image cropping for LLM | Suggests |
| `ellmer` | LLM extraction | Suggests |
| `validate` | Data quality rules | Suggests |
| `jsonlite` | JSON export | Suggests |
| `shiny` + `bslib` + `DT` | Standalone app and module | Suggests |
| `shinyFiles` | Filesystem browser in app | Suggests |
| `shinyAce` | YAML editor with syntax highlighting | Suggests |
| `shinyjs` | JS helpers in app | Suggests |
| `writexl` | Excel export | Suggests |
| `zip` | Zip-of-CSVs export | Suggests |
